A fast unified model for parsing and sentence understanding

### 2016-03-16 writing/hard_stack_paper/paper.tex
Low hanging fruit for length reduction.

Of these, it is easy to argue that the TreeRNN is the principled choice, since meaning in natural language sentences constructed incrementally according to a tree structure \citep{Partee84,Janssen97}. However, while these models have shown promise \citep{tai2015improved,li2015tree,bowman2015trees}, they have largely been overlooked in favor of sequence-based RNNs because of their incompatibility with batched computation and their reliance on external parsers.  Batched computation---performing synchronized computation across many examples at once---yields order-of-magnitude improvements in model run time, and is a crucial enabling technology in allowing neural networks to be trained efficiently on large datasets. Because TreeRNNs use a different model structure for each sentence, batched computation is impossible in standard implementations. In addition, standard TreeRNN models can only operate on sentences that have already been processed by a syntactic parser, which slows and complicates the use of these models at test time for most applications.

This paper introduces a new model to address both these issues: the Stack-augmented Parser-Interpreter Neural Network, or SPINN, shown in Figure~\ref{fig:m1-views}. SPINN executes the computations of a tree-structured model in a linearized sequence of operations, and can incorporate a neural network parser that determines the overall tree structure on-the-fly. This design improves upon the TreeRNN architecture in three ways:
\begin{itemize}
\item At test time, it can simultaneously parse and interpret unparsed sentences, removing the dependence on an external parser.
\item It supports batched computation for both parsed and unparsed sentences, supporting dramatic speedups over standard TreeRNNs.
\item It supports a novel tree-sequence hybrid mechanism for handling local context in sentence interpretation that yields substantial gains in accuracy over either pure sequence-based models or pure tree-based models.
\end{itemize}

We evaluate SPINN on the Stanford Natural Language Inference sentence understanding task \citep[SNLI,][]{snli:emnlp2015}, and find that it significantly outperforms other sentence encoding-based models, and that it yields speed increases of upwards of 10x over comparable TreeRNN models.

@@

The core SPINN model implements a sequential shift-reduce parser which operates over sentences. It is designed to produce a vector representation of the sentence as output, as opposed to a tree as in standard shift-reduce parsing. This relies on two key modifications to the shift-reduce algorithm:
\begin{itemize}
\item The parser's intermediate representations, including those of partial tree structures on the stack, are fixed-length vectors.
\item The \reduce~operation combines two such representations of trees on the stack into a larger tree using a version of the TreeLSTM neural network layer \citep{tai2015improved}.
\end{itemize}

@@

Our composition is a version of the TreeLSTM of \citet{tai2015improved}. The TreeLSTM layer, which generalizes the LSTM neural network layer to tree- rather than sequence-based inputs, and it shares with the LSTM the idea of representing intermediate states as a pair of a fast-changing state representation $\vec{h}$ and a slower-changing memory representation $\vec{c}$. It is formulated as:
\begin{gather}
\colvec{4}
    {\vec{i}}
    {\vec{f}_l}
    {\vec{f}_r}
    {\vec{o}}
= \sigma\left(
W_{\text{iffo}}
\colvec{3}
    {\vec{h}_s^1}
    {\vec{h}_s^2}
    {\vec{e}}
\right) \label{eqn:lstm1}
\\
\vec{g}
= \tanh\left(
W_{g}
\colvec{3}
    {\vec{h}_s^1}
    {\vec{h}_s^2}
    {\vec{e}}
\right) \label{eqn:lstm2}
\\
\vec{c} = \vec{f}_l \odot \vec{c}_s^{\,2} + \vec{f}_r \odot \vec{c}_s^{\,1} + \vec{i} \odot \vec{g}  
\\
\vec{h} = \vec{o} \odot \vec{c}
\end{gather}
The results of this function are the pair $\langle\vec{h}, \vec{c}\rangle$, which are placed back on the stack. The two input tree nodes popped off the stack are represented as the pairs $\langle\vec{h}^1_s, \vec{c}^{\,1}_s\rangle$ and $\langle\vec{h}^2_s, \vec{c}^{\,2}_s\rangle$. In addition, $\vec{e}$ is an optional input argument which is either the empty vector or a vector from an external source like the tracking LSTM (see Section~\ref{sec:tracking}). $\odot$ denotes the elementwise product. Each vector-valued variable listed here is of the same dimension $D$ except $\vec{e}$, which is of the independent dimension $D_t$.

@@

\paragraph{Creating a sentence pair classifier}

@@

In addition to the stack, the buffer, and the composition function, our full model includes an additional component: the tracking LSTM. The tracking LSTM is a simple low-dimensional sequence-based LSTM RNN that operates in tandem with the model, taking inputs from the buffer and stack at each step. It is meant to maintain a low-resolution summary of the portion of the sentence that has been processed so far, and it is used for two purposes: it supplies feature representations to the transition decision function, which allows the model to stand alone as a parser, and it additionally supplies a secondary input $\vec{e}$ (see Equations~\ref{eqn:lstm1}--\ref{eqn:lstm2}) to the composition function, allowing context information to leak into the construction of sentence meaning, and forming what is effectively a tree-sequence hybrid model.

@@

\todo{[JG] Include figure of post-order traversal sequence made by the tracking LSTM?}

@@

The model described so far is SPINN-NP (\ii{no parsing}), which uses sequences of \shift~and \reduce~operations from an external parser. To build the fully independent SPINN model, we simply add a single-layer neural network classifier that chooses which operation to perform at each step using the state of the tracking LSTM:
\begin{equation}
\vec{p}_{\text{trans}} = \text{softmax}(W_{\text{trans}}\vec{h}_{\text{track}})
\end{equation}
The resulting vector $\vec{p}_{\text{trans}}$ is a probability distribution over parsing decisions (i.e., predicting  \shift~or \reduce~at the current timestep). In the full SPINN model, we follow whichever transition has a higher predicted probability in this distribution. This prediction module is trained to mimic the decisions of an external parser. In particular, we use the transition sequences corresponding to the constituency parses included in the SNLI corpus. The classifier is trained using a cross-entropy objective:
\begin{equation}
  L_{\text{trans}} = - \log \vec{p}_{\text{trans}}[t^*]
\end{equation}
where $t^*$ is the correct transition specified by the gold parse, and $\vec{p}[\cdot]$ represents vector indexing.

@@

\paragraph{The full objective function} Our objective function combines a cross-entropy objective $\mathcal{L}_{\text{s}}$ for the SNLI classification task, a cross-entropy objective $\{\mathcal{L}_p^t, \mathcal{L}_h^t\}$ for each parsing decision for each of the sentence and the hypothesis sentence at each step $t$, and an L2 regularization term on the trained parameters. The terms are weighted using the tuned hyperparameters $\alpha$ and $\lambda$:
\begin{equation}
\begin{split}
\mathcal{L}_{\text{m}} = &\mathcal{L}_{\text{s}} + \alpha \sum_{t=0}^{T-1} (\mathcal{L}_{\text{p}}^{t} + \mathcal{L}_{\text{h}}^{t}) + \lambda \|\theta\|^2_2
\end{split}
\end{equation}
\paragraph{Preparing the data} At training time, SPINN requires both a transition sequence $\mathbf t$  and a token sequence $\mathbf x$ as its inputs for each sentence. The token sequence is simply the words in the sentence in order. $\mathbf t$ can be obtained from any constituency parse for the sentence by first converting that parse into an unlabeled binary parse, then linearizing it (with the usual in-order traversal), then taking each word token as a \shift~transition and each `)' as a \reduce~transition, as here:

@@

\paragraph{Software infrastructure} We will make source code available.

@@

Even though NLI is framed as a relatively simple three-way classification task, it is nonetheless an effective way of evaluating the ability of some model to extract broadly informative representations of sentence meaning. In order for a model to perform reliably well on NLI, it must be able to represent and reason with the core phenomena of natural language semantics, including quantification, coreference, scope, and several types of ambiguity.

@@

We evaluate four models on the SNLI task. Each uses 300D hidden states:
\begin{itemize}
\item A baseline LSTM model (similar to that of \citet{snli:emnlp2015}) that uses the same classifier architecture as our models, but encodes sentences using a single layer LSTM-RNN sequence model.
\item The minimal SPINN-NP-NT (\textit{no parsing, no tracking}), which uses transitions from an external parser and has no tracking LSTM to inform semantic composition. This model is equivalent to a TreeLSTM \citep{tai2015improved}.
\item SPINN-NP, which adds the tracking LSTM to guide composition, making it a hybrid tree-sequence model.
\item The full SPINN, which makes its own parsing decisions, and does not depend on an external parser at test time.
\end{itemize}

@@

\begin{figure}
\resizebox{\linewidth}{!}{
\begin{tikzpicture}
\begin{axis}[xmin=0, ymin=0, xlabel=Batch size, ylabel=Feedforward time (sec),
			 xtick={0,256,512,1024,2048}, legend style={at={(1.1, 0.98)},anchor=north east}]
  \addplot table [x={Batch size}, y=CPU, col sep=tab] {runtime.csv};
  \addlegendentry{CPU \citep{irsoy2014deep}}
  \addplot table [x={Batch size}, y=GPU, col sep=tab] {runtime.csv};
  \addlegendentry{Thin-stack GPU}
\end{axis}
\end{tikzpicture}
}
\caption{Runtime performance results with the thin-stack GPU algorithm. \todo{[JG] Shrink a bit.}}
\label{fig:speed}
\end{figure}

@@

Figure~\ref{fig:speed} compares the runtime performance of the two models computing representations on data from the Stanford Sentiment Treebank \todo{cite}. Each model is restricted to run on sentences of 30 tokens or fewer (for SPINN, this means we iterate over a sequence of 59 transitions). We fix the model dimensionality and word embedding dimensionality at 300 for all tests.

The CPU reference model is implemented in C++ and uses a single-core Eigen linear algebra backend. The implementation is a simplified version of the TreeRNN from \citet{irsoy2014deep}\footnote{The original code for this model is available at \url{https://github.com/oir/deep-recursive}. We plan to release our modified code.}, modified to exactly match the computation performed by our GPU model. We run the CPU performance test on a 2.20-GHz Intel Xeon processor with hyperthreading enabled. Our GPU model is implemented in C++ and CUDA, and will be released with the rest of our code. We test its performance on an NVIDIA Titan X GPU.

@@

This paper shows that:
\begin{itemize}
\item SPINN-NP-NT is exactly equivalent to a tree-structured network in what it computes, but is 10--20 times faster than a standard tree network at inference time.
\item While the use of tree-structure alone does not offer a substantial improvement over an LSTM baseline, our base tree-sequence hybrid model (SPINN-NP) substantially outperforms that baseline, suggesting that our model is able to successfully exploit tree structure, while using sequential context information to better resolve ambiguity.
\item It is possible to remove the dependence on an external parser from our model and replace it with a fast model-internal parser (as in the full SPINN) with only a small decrease in model performance.
\end{itemize}

@@

For a more ambitious goal, we expect that it should be possible to implement a variant of SPINN on top of a modified stack data structure with differentiable \textsc{push} and \textsc{pop} operations (possibly following \citet{grefenstette2015learning,joulin2015inferring}). This would make it possible for the model to learn to parse using guidance from the semantic representation objective, essentially allowing learn to produce parses that are, in aggregate, better suited to supporting semantic interpretation than those supplied in the training data.

>>>>

Of these, the TreeRNN is the principled choice, since meaning in natural language sentences constructed incrementally according to a tree structure \citep{Partee84,Janssen97}. However, while these models have shown promise \citep{tai2015improved,li2015tree,bowman2015trees}, they have largely been overlooked in favor of sequence-based RNNs because of their incompatibility with batched computation and their reliance on external parsers.  Batched computation---performing synchronized computation across many examples at once---yields order-of-magnitude improvements in model run time, and is a crucial enabling technology in allowing neural networks to be trained efficiently on large datasets. Because TreeRNNs use a different model structure for each sentence, batched computation is impossible in standard implementations. In addition, standard TreeRNN models can only operate on sentences that have already been processed by a syntactic parser, which slows and complicates the use of these models at test time for most applications.

This paper introduces a new model to address both these issues: the Stack-augmented Parser-Interpreter Neural Network, or SPINN, shown in Figure~\ref{fig:m1-views}. SPINN executes the computations of a tree-structured model in a linearized sequence of operations, and can incorporate a neural network parser that determines the overall tree structure on-the-fly. This design improves upon the TreeRNN architecture in three ways: At test time, it can simultaneously parse and interpret unparsed sentences, removing the dependence on an external parser. In addition, it supports batched computation for both parsed and unparsed sentences, supporting dramatic speedups over standard TreeRNNs. Finally, it supports a novel tree-sequence hybrid mechanism for handling local context in sentence interpretation that yields substantial gains in accuracy over either pure sequence-based models or pure tree-based models.

We evaluate SPINN on the Stanford Natural Language Inference entailment task \citep[SNLI,][]{snli:emnlp2015}, and find that it significantly outperforms other sentence encoding-based models, and that it yields speed increases of over 10x over a standard TreeRNN implementation.

@@

The core SPINN model implements a sequential shift-reduce parser which operates over sentences. It is designed to produce a vector representation of the sentence as output, as opposed to a tree as in standard shift-reduce parsing. This relies on two key modifications to the shift-reduce algorithm: First, the parser's intermediate representations, including those of partial tree structures on the stack, are fixed-length vectors. In addition, the \reduce~operation combines two such representations of trees on the stack into a larger tree using a version of the TreeLSTM neural network layer \citep{tai2015improved}.

@@

The TreeLSTM composition function generalizes the LSTM neural network layer to tree- rather than sequence-based inputs, and it shares with the LSTM the idea of representing intermediate states as a pair of a fast-changing state representation $\vec{h}$ and a slower-changing memory representation $\vec{c}$. Our version is formulated as:
\begin{gather}
\colvec{5}
    {\vec{i}}
    {\vec{f}_l}
    {\vec{f}_r}
    {\vec{o}}
    {\vec{g}}
= \colvec{5}
    {\sigma}
    {\sigma}
    {\sigma}
    {\sigma}
    {\text{tanh}}
\left(
W_{\text{iffo}}
\colvec{3}
    {\vec{h}_s^1}
    {\vec{h}_s^2}
    {\vec{e}}
\right) \label{eqn:lstm1}
\\
\vec{c} = \vec{f}_l \odot \vec{c}_s^{\,2} + \vec{f}_r \odot \vec{c}_s^{\,1} + \vec{i} \odot \vec{g}  
\\
\vec{h} = \vec{o} \odot \vec{c}
\end{gather}
$\sigma$ is the sigmoid activation function. The results of this function are the pair $\langle\vec{h}, \vec{c}\rangle$, which are placed back on the stack. The two input tree nodes popped off the stack are represented as the pairs $\langle\vec{h}^1_s, \vec{c}^{\,1}_s\rangle$ and $\langle\vec{h}^2_s, \vec{c}^{\,2}_s\rangle$. In addition, $\vec{e}$ is an optional input argument which is either the empty vector or a vector from an external source like the tracking LSTM (see Section~\ref{sec:tracking}). $\odot$ denotes the elementwise product. Each vector-valued variable listed here is of the same dimension $D$ except $\vec{e}$, which is of the independent dimension $D_t$.

@@

\paragraph{Creating a sentence pair classifier} \label{sec:classifier}

@@

In addition to the stack, the buffer, and the composition function, our full model includes an additional component: the tracking LSTM. The tracking LSTM is a simple low-dimensional sequence-based LSTM RNN that operates in tandem with the model, taking inputs from the buffer and stack at each step. It is meant to maintain a low-resolution summary of the portion of the sentence that has been processed so far, and it is used for two purposes: it supplies feature representations to the transition decision function, which allows the model to stand alone as a parser, and it additionally supplies a secondary input $\vec{e}$ (see Equation~\ref{eqn:lstm1}) to the composition function, allowing context information to leak into the construction of sentence meaning, and forming what is effectively a tree-sequence hybrid model.

@@

\todo{[SB] Shorten the previous paras.}

@@

The model described so far is SPINN-NP (\ii{no parsing}), which uses sequences of \shift~and \reduce~operations from an external parser. To build the fully independent SPINN model, we simply add a single-layer neural network classifier that chooses which operation to perform at each step using the state of the tracking LSTM:
\begin{equation}
\vec{p}_{\text{t}} = \text{softmax}(W_{\text{transition}}\vec{h}_{\text{tracking}})
\end{equation}
The resulting vector $\vec{p}_{\text{t}}$ is a probability distribution over parsing decisions (i.e., predicting  \shift~or \reduce~at the current timestep). In the full SPINN model, we follow whichever transition has a higher predicted probability in this distribution. This prediction module is trained to mimic the decisions of an external parser. In particular, we use the transition sequences corresponding to the constituency parses included in the SNLI corpus.

@@

\paragraph{The objective function} Our objective function combines a cross-entropy objective $\mathcal{L}_{\text{s}}$ for the SNLI classification task, a cross-entropy objective $\{\mathcal{L}_p^t, \mathcal{L}_h^t\}$ for each parsing decision for each of the sentence and the hypothesis sentence at each step $t$, and an L2 regularization term on the trained parameters. The terms are weighted using the tuned hyperparameters $\alpha$ and $\lambda$:
\begin{equation}
\begin{split}
\mathcal{L}_{\text{m}} = &\mathcal{L}_{\text{s}} + \alpha \sum_{t=0}^{T-1} (\mathcal{L}_{\text{p}}^{t} + \mathcal{L}_{\text{h}}^{t}) + \lambda \|\theta\|^2_2
\end{split}
\end{equation}
\paragraph{Preparing the data} At training time, SPINN requires both a transition sequence $\mathbf t$  and a token sequence $\mathbf x$ as its inputs for each sentence. The token sequence is simply the words in the sentence in order. $\mathbf t$ can be obtained from any constituency parse for the sentence by first converting that parse into an unlabeled binary parse, then linearizing it (with the usual in-order traversal), then taking each word token as a \shift~transition and each `)' as a \reduce~transition, as here:

@@

@@

Although NLI is framed as a simple three-way classification task, it is nonetheless an effective way of evaluating the ability of some model to extract broadly informative representations of sentence meaning. In order for a model to perform reliably well on NLI, it must be able to represent and reason with the core phenomena of natural language semantics, including quantification, coreference, scope, and several types of ambiguity.

@@

We evaluate four models on the SNLI task. Each uses the sentence pair classifier architecture described in Section~\ref{sec:classifier}, and they differ only in the function computing the sentence encodings. First, a single-layer LSTM RNN \citep[similar to that of][]{snli:emnlp2015} serves as a baseline. Next, the minimal SPINN-NP-NT model (exactly equivalent to a TreeLSTM) introduces the SPINN model design. SPINN-NP adds the tracking LSTM to that design. Finally, the full SPINN adds the integrated parser.

@@

\begin{figure}
\resizebox{0.85\linewidth}{!}{
\begin{tikzpicture}
\begin{axis}[xmin=0, ymin=0, xlabel=Batch size, ylabel=Feedforward time (sec),
			 xtick={0,256,512,1024,2048}, legend style={at={(1.1, 0.98)},anchor=north east}]
  \addplot table [x={Batch size}, y=CPU, col sep=tab] {runtime.csv};
  \addlegendentry{CPU \citep{irsoy2014deep}}
  \addplot table [x={Batch size}, y=GPU, col sep=tab] {runtime.csv};
  \addlegendentry{Thin-stack GPU}
\end{axis}
\end{tikzpicture}
}
\caption{Runtime performance results with the thin-stack GPU algorithm.}}
\label{fig:speed}
\end{figure}

@@

Figure~\ref{fig:speed} compares the runtime performance of the two models computing representations on data from the Stanford Sentiment Treebank \citep{socher2013recursive}. Each model is restricted to run on sentences of 30 tokens or fewer (for SPINN, this means we iterate over a sequence of 59 transitions). We fix the model dimension and the word embedding dimension at 300.

The CPU reference model is implemented in C++ and uses a single-core Eigen linear algebra backend. The implementation is a simplified version of the TreeRNN from \citet{irsoy2014deep}\footnote{The original code for this model is available at \url{https://github.com/oir/deep-recursive}. We plan to release our modified code alongside the full SPINN source code.}, modified to exactly match the computation performed by our GPU model. We run the CPU performance test on a 2.20-GHz Intel Xeon processor with hyperthreading enabled. Our GPU model is implemented in C++ and CUDA, and will be released with the rest of our code. We test its performance on an NVIDIA Titan X GPU.

@@

In this paper, we first introduce a model architecture (SPINN-NP-NT) that is precisely equivalent to a TreeLSTM, but an order of magnitude faster at test time. Next, we expand that architecture into a tree-sequence hybrid model (SPINN-NP), and show that this yields significant gains on the SNLI entailment task. Finally, we show that it is possible exploit the strengths of this model without the need for an external parser by integrating a fast parser into the model itself, as in the full SPINN.

@@

For a more ambitious goal, we expect that it should be possible to implement a variant of SPINN on top of a modified stack data structure with differentiable \textsc{push} and \textsc{pop} operations \citep[as in][]{grefenstette2015learning,joulin2015inferring}. This would make it possible for the model to learn to parse using guidance from the semantic representation objective, essentially allowing learn to produce parses that are, in aggregate, better suited to supporting semantic interpretation than those supplied in the training data.
